---
title: "biomarkers_single_patient"
output: html_document
---

```{r}
#packages and custom functions
source("functions.R")
pacman::p_load(tidybayes, bayesplot, posterior)
```

# Bayesian Analysis of Patient id 12

This script loads in a Bayesian fit of the lokta volterra model to
patient 12 from the Bruchovsky cohort. The model is fitted using Stan -
basically just a wrapper around C++ that runs MCMC. The model has fairly
tight priors (see corresponding script. The following parameters were
allowed to vary:

- cost

- turnover: cell death rate

- rFrac: fraction of resitant cells at t=0)

- n0: initial cell density

- K: carrying capacity of tomour

- rS: proliferation rate of sensitive cells

- sigma: noise parameter

```{r}
#load in the patient and print summary of fitted parameters
patient_id = 12

target_dir = here::here("fits/free_rs_K_full_history")
fit_path <- make_fit_path(patient_id = patient_id, dir = target_dir)
fit <- readRDS(fit_path)

#show the summary
fit$summary(c("turnover", "cost", "n0", "rFrac", "K", "rS", "sigma"))
```

This is a summary of the 3000 parameter estimates drawn according to the
probability given the data, priors and the ODE model assuming that
normalized PSA at time T correspond to Gaussian noise around the total
cell population (S+R) at the same timepoint. MCMC samples can be
accessed (with a bit of wrangling)

```{r}
#get draws in tidy format of free params
draws <- fit$draws(c("cost", "turnover", "n0", "rFrac", "sigma", "rS", "K")) |>                                       
    spread_draws(cost, turnover, n0, rFrac, sigma, K, rS) |>                                                          
    select(c(-.chain, -.iteration))  #rename starting states 

#show draws table
draws

```

We can use this to plot a density function over the parameters. For
example *turnover*

```{r}
ggplot(draws, aes(x=turnover)) +
  geom_density(fill = "orange") +
  theme_classic() +
  ggtitle(paste0("Turnover for Patient ID ", patient_id)) + 
  xlim(-.1, 1)

```

## Adding Biomarkers

For each parameter draw we can define functions that take a draw in an
compute a value based on the draw to propogate the uncertinaty. For
example, we can calculate the biomarkers AT score and eTTP for each
draw.

```{r}
#rebuild draws object with all variables that we need
draws <- fit$draws(c("cost", "turnover", "n0", "rFrac", "sigma", "rS", "K", "rR", "dR", "starting_state")) |>
    #wrangle stan output
    spread_draws(cost, turnover, n0, rFrac, sigma, rR, dR, K, rS, starting_state[j]) |>                                                          
    pivot_wider(names_from = j, values_from = starting_state) |> #have to wrangle a bit to get starting states                                                           
    rename(S0 = "1", R0 = "2") |>   
  
    #calculate biomarkers 
    mutate(AT_score = calc_delta_benefit(kappa=.2, dR = dR, K=K, rR = rR, R0 = R0), #check with kit whether kappa=.02 makes sense
           eTTP = predict_atx_ttp(tau = 30, rS = rS, dR = dR, dD = 1.5, dS = dR, K = K, rR = rR, R0 = R0, S0 = S0, n0 = n0)) |> 
  select(c(-.iteration, -.chain)) |> 
  relocate(c(.draw, AT_score, eTTP))

#show
draws
```

We can plot the densities of the samples

```{r}
#plotting
ggplot(draws, aes(x=AT_score)) +
  geom_density(fill = "skyblue") +
  theme_classic() +
  ggtitle(paste0("AT Score Patient ID ", patient_id), "Full Treatment History") +
  xlim(0, 1) 
```

We also get samples of intermediate / determined parameters such as the
inferred cell populations at each step. Here we overlay each of the 3000
trajectories in cell populations.

```{r}
#rebuild the data that was fed into stan to get measurement times back
patient_data <- load_bruchowsky_patients(patient_id)[[as.character(patient_id)]]
obs <- patient_data |> 
  filter(!is.na(PSA))

#wrangle parameter draws. Each draw is an entire trajectory
cell_population_draws <- fit$draws("cell_populations") |>
    spread_draws(cell_populations[i, k]) |>
    #pivot_wider(names_from = k, values_from = cell_populations, names_prefix = "pop") |>
    mutate(cell_type = if_else(k==1, "S", "R"),
           Time = obs$Time[i]) |> 
    rename(obs_idx = i) 


#plot it
ggplot(cell_population_draws, aes(x = Time, y = cell_populations, color = cell_type)) +
  geom_point(alpha = .01) + #draw inferred pops
  #geom_point(data=obs, aes(x = Time, y= PSA), shape = "x", color ="black", size = 2) + #overlay psa
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 3)))  +
  theme_classic() +
  xlab("Time (Days) ") +
  ylab("Cell Density") +
  ggtitle("Inferred Cell Densities")
```

# Using only data from the first cycle

The following shows inference when only the first treatment cycle
(CycleId = 1) is used.

- cost

- turnover:

- rFrac: fraction of resistant cells at t=0

- n0: initial cell density

- K: carrying capacity of tumour

- rS: proliferation rate of sensitive cells

- sigma: noise parameter

Read in the first cycle fits of all five patients and calculate the AT
score for every draw.

```{r}

#load in first cycle fits
patient_ids <- c(12, 20, 25, 78, 85)
target_dir <- here::here("fits/free_rs_K_first_cycle")   # or free_rs_K_full_history

for (id in patient_ids) {
  fit <- readRDS(make_fit_path(patient_id = id, dir = target_dir))

  at_draws <- fit$draws(c("rR", "dR", "K", "starting_state")) |>
    spread_draws(rR, dR, K, starting_state[j]) |>
    pivot_wider(names_from = j, values_from = starting_state) |>
    rename(S0 = "1", R0 = "2") |>
    mutate(PatientId = id,
           AT_score = calc_delta_benefit(kappa = .2, dR = dR, K = K, rR = rR, R0 = R0)) |>
    select(PatientId, .draw, AT_score)
  
  plot <- ggplot(filter(at_draws, is.finite(AT_score)), aes(x = AT_score)) +
    geom_density(fill = "skyblue") +
    ggtitle(paste0("First cycle AT score for patient ", id)) +
    theme_classic() +
    xlim(-5,5)
  
  print(plot)
}


```

Looks very uncertain. R is also complaining that AT score could be
computed for approx 280 rows computed, so this a biased picture.

We can also check how much the posteriors moved from the priors.

```{r}
#clauded up a function plot posteriors versus priors
plot_prior_posterior <- function(patient_id, n = 4000) {
  first_fit <- readRDS(make_fit_path(patient_id, here::here("fits/free_rs_K_first_cycle")))
  full_fit  <- readRDS(make_fit_path(patient_id, here::here("fits/free_rs_K_full_history")))
  
  prior_draws <- tibble(
    cost     = rbeta(n, 5, 10),
    turnover = rbeta(n, 2, 2),
    n0       = rbeta(n, 2, 2),
    rFrac    = rbeta(n, 2, 30),
    K        = rlnorm(n, log(2), .4),
    rS       = rlnorm(n, log(0.019), .25),
    sigma    = rlnorm(n, -2, 1)
  ) |>
    filter(K > n0)
  
  params <- names(prior_draws)
  get_post <- function(fit) {
    fit$draws(params, format = "draws_df") |> as_tibble() |> select(all_of(params))
  }
  
  bind_rows(prior            = prior_draws,
            `first cycle`    = get_post(first_fit),
            `full history`   = get_post(full_fit),
            .id = "dist") |>
    mutate(dist = factor(dist, c("prior", "first cycle", "full history"))) |>
    pivot_longer(-dist, names_to = "parameter") |>
    ggplot(aes(value, fill = dist)) +
    geom_density(alpha = .5, colour = NA) +
    facet_wrap(~ parameter, scales = "free") +
    scale_fill_manual(values = c("prior" = "grey70", "first cycle" = "#2a78d6", "full history" = "#eb6834")) +
    theme_classic() +
    labs(title = paste("Prior and posteriors, patient", patient_id), fill = NULL)
}

for (id in c(12, 20, 25, 78, 85)) {
  print(plot_prior_posterior(id))
}
```

## **What would be a meaningful contribution?**

Would like to implement the hierarchical model that you suggest to
estimate population level priors that could be used for first cycle
fits, and investigate how accurate inference that would allow us to make
about biomarkers. Building something that clinicians could use as
cut-off to select who gets what treatment. I'm guessing this would be
maximally informed by fitting finding some population level prior. For
example a hierarchical model of the Bruchovsky data.

# Current issues:

### eTTP and expected mean daily dose

Thought, I had gotten the eTTP right yesterday, but checked against Fig
1 in "Mathematical Biomaker" paper had completely different estimates
than the paper. *Would like to know what functions correspond to what
measures, and would be most meaningful to report going forward.*

### Sampling

At least two things can go wrong with the sampler

- The easy to spot problem: The sampler breaks down in some areas.

- The difficult to spot problem: sampler is not converging. There are
  some easy to use metrics, but to my understanding the gold standard is
  running thousands of forward simulations and investigating how the
  model performs on these. Haven't looked much at this yet, since it
  requires fitting the model thousands of times and its currently
  breaking down and slow.

```{r}
patient_ids <- c(12, 20, 25, 78, 85)

for (id in patient_ids){

  #read in the fit
  fit_path <- make_fit_path(patient_id = id, dir = target_dir)
  fit <- readRDS(fit_path)

  #get fit summary
  print(paste0("Patient ID: ", id))
  fit_summary <- as.data.frame(fit$summary(c("cost", "turnover", "rFrac","n0", "K", "rS", "sigma"))) |> 
    mutate(across(where(is.numeric), ~ round(.x, 3)))
  print(fit_summary)
  print(fit$diagnostic_summary())

  #draw pairs
  print(mcmc_pairs(fit$draws(c("cost", "turnover", "n0", "rFrac","K", "rS", "sigma")), np = nuts_params(fit))
  )
}
```

### 
